---
title: "Estimation of exposure level using inversed distance weighting"
author: "Shuangjia Xue"
date: "2022-11-15"
output: 
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    number_sections: yes
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE, echo=F}
library(here)
library(sp)
library(gstat)
library(stringr)
library(ggmap)
library(plyr)
```

### IDW sample https://rpubs.com/hungle510/202761

### input data
```{r}
setwd('C:/Air pollution/Air-pollution-working-with-Nouraie/data')

SO2 <- read.csv("SO2_monthly_2000_2022.csv")
SO2[1:6,1:6]
```


### create a known exposure level dataframe (coordiates, concentration)
```{r}
lat <- str_split_fixed(SO2$monitor_id,',', 2)[,1] %>% as.numeric()
long <- str_split_fixed(SO2$monitor_id,',', 2)[,2] %>% as.numeric()
SO2_2000_1 <- SO2$X2000.1 

knowndt = data.frame(lat,long,SO2_2000_1)

coordinates(knowndt)  <- ~ lat + long
```


### five random subjects 
```{r}
X1 = c(40.44343,42.01629,40.49899,26.79027,44.94620); Y1 = c(-79.95497,-88.23463,-74.68470,80.06925,-91.40844)

unknowndt = data.frame(X1,Y1)

coordinates(unknowndt)  <- ~ X1 + Y1
```


###  Inverse Distance Weight (IDW) for air pollution estimation
## weights are proportional to the inverse of the distance 
#(between the data point and the prediction location) 
#raised to the power value p. As a result, as the distance increases, 
#the weights decrease rapidly. 

```{r}
knowndt1 <- knowndt[!is.na(knowndt$SO2_2000_1),]  ## remove NA

idwmodel = idw(SO2_2000_1 ~1, knowndt1,unknowndt,
               maxdist = Inf)
idwmodel$var1.pred 
```


## create the initial summary dataframe

```{r}
summary_SO2 <- data.frame(1:5,idwmodel$var1.pred) 
colnames(summary_SO2) <-c('ID','2000/1')

head(summary_SO2)
```


#### running loop for remaining date
```{r}
for (i in 3:ncol(SO2)) {
  date <- colnames(SO2)[i]
  SO2_tmp <- SO2[,i]
  knowndt_tmp <- data.frame(lat,long,SO2_tmp)
  coordinates(knowndt_tmp)  <- ~ lat + long
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$SO2_tmp),]
  idwmodel1 = idw(SO2_tmp ~1, known_na_omit,unknowndt,
                  maxdist = Inf)
  summary_SO2[,i] <- idwmodel1$var1.pred
  colnames(summary_SO2)[i] <- date
  
} 


```



## change date format

```{r}

summary <- summary_SO2
colnames(summary) <- gsub('X','',colnames(summary))
colnames(summary) <- gsub('\\.','/',colnames(summary))

```



### modify the date
```{r}

dates <- paste0(colnames(summary),"/01") ## add day1 to month and year

## remove first two digits
dates1 <- substr(dates, 3, nchar(dates))

### use as.Date function
betterDates <- as.Date(dates1, "%y/%m/%d")
colnames(summary)<-betterDates
```


### define date of diagnosis

```{r}
diag <- c('2007-03-30','2008-07-31','2010-05-20','2012-03-31','2009-01-02')

nrow(summary)
summary <- summary[,-1]
```


### run a loop for every patinet to remove data after their diagnosis

```{r}
n <- 5
datalist <- list()
datalist <- vector("list", length = n)

for (i in 1:n) {
  datalist[[i]] <- summary[i,][colnames(summary) <= diag[i]] # add it to your list
}


```


```{r}
SO2_summary <- do.call(rbind.fill, datalist)

```