---
title: "Variogram for major air pollutants and kriging estimation"
author: "Shuangjia (Nicole) Xue"
date: "2022-11-12"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_sections: yes
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE, echo=F}
library(sp)
library(bookdown)
library(gridExtra)
library(gstat)
library(stringr)
library(plyr)
library(gstat)
library(zipcodeR)
library(ggplot2)
library(maptools)
library(scales) 
library(magrittr)
```

### reference https://rpubs.com/nabilabd/118172


### input data
```{r}
setwd('C:/Air pollution/Air-pollution-working-with-Nouraie/data')

PM2.5 <- read.csv("PM2.5_monthly_2000_2022.csv")
PM10 <- read.csv("PM10_monthly_2000_2022.csv")
O3 <- read.csv("O3_monthly_2000_2022.csv")
CO <- read.csv("CO_monthly_2000_2022.csv")
SO2 <- read.csv("SO2_monthly_2000_2022.csv")
NO2 <- read.csv("NO2_monthly_2000_2022.csv")
```


### build coordinates
```{r}
#PM2.5
PM2.5$long <- str_split_fixed(PM2.5$monitor_id,',', 2)[,2] %>% as.numeric()
PM2.5$lat <- str_split_fixed(PM2.5$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(PM2.5) <- ~ long + lat

#PM10
PM10$long <- str_split_fixed(PM10$monitor_id,',', 2)[,2] %>% as.numeric()
PM10$lat <- str_split_fixed(PM10$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(PM10) <- ~ long + lat

#SO2
SO2$long <- str_split_fixed(SO2$monitor_id,',', 2)[,2] %>% as.numeric()
SO2$lat <- str_split_fixed(SO2$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(SO2) <- ~ long + lat

#CO
CO$long <- str_split_fixed(CO$monitor_id,',', 2)[,2] %>% as.numeric()
CO$lat <- str_split_fixed(CO$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(CO) <- ~ long + lat

#NO2
NO2$long <- str_split_fixed(NO2$monitor_id,',', 2)[,2] %>% as.numeric()
NO2$lat <- str_split_fixed(NO2$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(NO2) <- ~ long + lat

#O3
O3$long <- str_split_fixed(O3$monitor_id,',', 2)[,2] %>% as.numeric()
O3$lat <- str_split_fixed(O3$monitor_id,',', 2)[,1] %>% as.numeric()
coordinates(O3) <- ~ long + lat

```

# monitor locations
```{r}
P1<-plot(PM2.5)
P2<-plot(PM10)
P3<-plot(SO2)
P4<-plot(CO)
P5<-plot(NO2)
P6<-plot(O3)

grid.arrange(P1,P2,P3,P4,P5,P6, ncol = 2)

```

# plot with exposure level for Jan 2000
```{r}
p1<-PM2.5 %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

p2<-PM10 %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

p3<-SO2 %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

p4<-CO %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

p5<-NO2 %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

p6<-O3 %>% as.data.frame %>% 
  ggplot(aes(long, lat)) + geom_point(aes(size = X2000.1), color="blue", alpha=3/4) +  coord_equal() + theme_bw()

grid.arrange(p1,p2,p3,p4,p5,p6, ncol = 2)

```

# variogram for PM2.5 in Jan 2000

```{r}

PM2.5_NAomit <- PM2.5[which(PM2.5$X2000.1 != 'NA'),] # remove NA
pm.vgm.pm2.5 <- variogram(X2000.1~1, data = PM2.5_NAomit) # calculates sample variogram values 
pm.fit.pm2.5 <- fit.variogram(pm.vgm.pm2.5, vgm(1,"Gau",100,1)) # fit model

plot(pm.vgm.pm2.5, pm.fit.pm2.5)

```


# PM10
```{r}

PM10_NAomit <- PM10[which(PM10$X2000.1 != 'NA'),] # remove NA
pm.vgm.pm10 <- variogram(X2000.1~1, data = PM10_NAomit) # calculates sample variogram values 
pm.fit.pm10 <- fit.variogram(pm.vgm.pm10, vgm(1,"Gau",100,1)) # fit model

plot(pm.vgm.pm10, pm.fit.pm10)

```

#SO2

```{r}

SO2_NAomit <- SO2[which(SO2$X2000.1 != 'NA'),] # remove NA
TheVariogram.SO2 <- variogram(X2000.1~1, data=SO2_NAomit) # calculates sample variogram values 
TheVariogramModel.SO2 <- vgm(psill=0.15, model="Gau", nugget=0.0001, range=5)
FittedModel.SO2 <- fit.variogram(TheVariogram.SO2, model=TheVariogramModel.SO2)    
plot(TheVariogram.SO2, model=FittedModel.SO2)

```

#CO

```{r}

CO_NAomit <- CO[which(CO$X2000.1 != 'NA'),] # remove NA
TheVariogram.CO <- variogram(X2000.1~1, data=CO_NAomit) # calculates sample variogram values 
TheVariogramModel.CO <- vgm(psill=0.15, model="Gau", nugget=0.0001, range=5)
FittedModel.CO <- fit.variogram(TheVariogram.CO, model=TheVariogramModel.CO)    
plot(TheVariogram.CO, model=FittedModel.CO)

```

#NO2

```{r}

NO2_NAomit <- NO2[which(NO2$X2000.1 != 'NA'),] # remove NA
TheVariogram.NO2 <- variogram(X2000.1~1, data=NO2_NAomit) # calculates sample variogram values 
TheVariogramModel.NO2 <- vgm(psill=0.15, model="Gau", nugget=0.0001, range=5)
FittedModel.NO2 <- fit.variogram(TheVariogram.NO2, model=TheVariogramModel.NO2)    
plot(TheVariogram.NO2, model=FittedModel.NO2)

```

#O3

```{r}

O3_NAomit <- O3[which(O3$X2000.1 != 'NA'),] # remove NA
TheVariogram.O3 <- variogram(X2000.1~1, data=O3_NAomit) # calculates sample variogram values 
TheVariogramModel.O3 <- vgm(psill=0.15, model="Gau", nugget=0.0001, range=5)
FittedModel.O3 <- fit.variogram(TheVariogram.O3, model=TheVariogramModel.O3)    
plot(TheVariogram.O3, model=FittedModel.O3)

```



# ten zip code from Allegeny County
```{r, message=FALSE, warning=FALSE, echo=F}
zipcodes <- c('15213','15049','15122','15102','15203','15014',
              '15024','15104','15129','15006')
```


# find longitude and latitude for each zip code using zipcodeR
```{r, message=FALSE, warning=FALSE, echo=F}
n <- length(zipcodes)
coord <- data.frame()
for (i in 1:n) {
  coord <- rbind(coord, geocode_zip(zipcodes[i]))
}
coord
unknowndt <- coord[,-1]
head(unknowndt)
coordinates(unknowndt)  <- ~ lng + lat
```

# using Kriging to estimate exposure level for each patients and loop through all the months

## input the data again to remove coordinates
```{r,warning=FALSE}
setwd('C:/Air pollution/Air-pollution-working-with-Nouraie/data')
PM2.5 <- read.csv("PM2.5_monthly_2000_2022.csv")
PM10 <- read.csv("PM10_monthly_2000_2022.csv")
O3 <- read.csv("O3_monthly_2000_2022.csv")
CO <- read.csv("CO_monthly_2000_2022.csv")
SO2 <- read.csv("SO2_monthly_2000_2022.csv")
NO2 <- read.csv("NO2_monthly_2000_2022.csv")
```


# PM2.5
```{r, message=FALSE, warning=FALSE, echo=F}
# PM2.5
PM2.5_Krig <- data.frame(zipcodes) # can change to subject IDs 

long <- str_split_fixed(PM2.5$monitor_id,',', 2)[,2] %>% as.numeric()
lat <- str_split_fixed(PM2.5$monitor_id,',', 2)[,1] %>% as.numeric()



for (i in 2:ncol(PM2.5)) {
  date <- colnames(PM2.5)[i]
  PM2.5_tmp <- PM2.5[,i]
  knowndt_tmp <- data.frame(lat,long,PM2.5_tmp)
  coordinates(knowndt_tmp)  <- ~ long + lat
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$PM2.5_tmp),]
  pm2.5.kriged <-krige(PM2.5_tmp ~ 1, known_na_omit, unknowndt, model = pm.fit.pm2.5, maxdist=10, nmax=50)
  PM2.5_Krig[,i] <- pm2.5.kriged$var1.pred
  colnames(PM2.5_Krig)[i] <- date
  
} 

PM2.5_Krig[1:6,1:6]
```




# PM10
```{r, message=FALSE, warning=FALSE, echo=F}
PM10_Krig <- data.frame(zipcodes) # can change to subject IDs 

long <- str_split_fixed(PM10$monitor_id,',', 2)[,2] %>% as.numeric()
lat <- str_split_fixed(PM10$monitor_id,',', 2)[,1] %>% as.numeric()



for (i in 2:ncol(PM10)) {
  date <- colnames(PM10)[i]
  PM10_tmp <- PM10[,i]
  knowndt_tmp <- data.frame(lat,long,PM10_tmp)
  coordinates(knowndt_tmp)  <- ~ long + lat
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$PM10_tmp),]
  PM10.kriged <-krige(PM10_tmp ~ 1, known_na_omit, unknowndt, model = pm.fit.pm10, maxdist=10, nmax=50)
  PM10_Krig[,i] <- PM10.kriged$var1.pred
  colnames(PM10_Krig)[i] <- date
  
} 

PM10_Krig[1:6,1:6]
```



# CO
```{r, message=FALSE, warning=FALSE, echo=F}
CO_Krig <- data.frame(zipcodes) # can change to subject IDs 

long <- str_split_fixed(CO$monitor_id,',', 2)[,2] %>% as.numeric()
lat <- str_split_fixed(CO$monitor_id,',', 2)[,1] %>% as.numeric()



for (i in 2:ncol(CO)) {
  date <- colnames(CO)[i]
  CO_tmp <- CO[,i]
  knowndt_tmp <- data.frame(lat,long,CO_tmp)
  coordinates(knowndt_tmp)  <- ~ long + lat
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$CO_tmp),]
  CO.kriged <-krige(CO_tmp ~ 1, known_na_omit, unknowndt, model = FittedModel.CO, maxdist=10, nmax=50)
  CO_Krig[,i] <- CO.kriged$var1.pred
  colnames(CO_Krig)[i] <- date
  
} 

CO_Krig[1:6,1:6]
```


# NO2
```{r, message=FALSE, warning=FALSE, echo=F}
NO2_Krig <- data.frame(zipcodes) # can change to subject IDs 

long <- str_split_fixed(NO2$monitor_id,',', 2)[,2] %>% as.numeric()
lat <- str_split_fixed(NO2$monitor_id,',', 2)[,1] %>% as.numeric()



for (i in 2:ncol(NO2)) {
  date <- colnames(NO2)[i]
  NO2_tmp <- NO2[,i]
  knowndt_tmp <- data.frame(lat,long,NO2_tmp)
  coordinates(knowndt_tmp)  <- ~ long + lat
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$NO2_tmp),]
  NO2.kriged <-krige(NO2_tmp ~ 1, known_na_omit, unknowndt, model = FittedModel.NO2, maxdist=10, nmax=50)
  NO2_Krig[,i] <- NO2.kriged$var1.pred
  colnames(NO2_Krig)[i] <- date
  
} 

NO2_Krig[1:6,1:6]
```




# O3
```{r, message=FALSE, warning=FALSE, echo=F}
O3_Krig <- data.frame(zipcodes) # can change to subject IDs 

long <- str_split_fixed(O3$monitor_id,',', 2)[,2] %>% as.numeric()
lat <- str_split_fixed(O3$monitor_id,',', 2)[,1] %>% as.numeric()



for (i in 2:ncol(O3)) {
  date <- colnames(O3)[i]
  O3_tmp <- O3[,i]
  knowndt_tmp <- data.frame(lat,long,O3_tmp)
  coordinates(knowndt_tmp)  <- ~ long + lat
  known_na_omit <- knowndt_tmp[!is.na(knowndt_tmp$O3_tmp),]
  O3.kriged <-krige(O3_tmp ~ 1, known_na_omit, unknowndt, model = FittedModel.O3, maxdist=10, nmax=50)
  O3_Krig[,i] <- O3.kriged$var1.pred
  colnames(O3_Krig)[i] <- date
  
} 

O3_Krig[1:6,1:6]
```
